Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2020-12-14 GWLDCI.tex
final changes discussed in meeting for paper 1

%*****************
% Annotations
\usepackage{soul}
\usepackage[colorinlistoftodos,textsize=footnotesize]{todonotes}
\newcommand{\hlfix}[2]{\texthl{#1}\todo{#2}}
\newcommand{\hlnew}[2]{\texthl{#1}\todo[color=green!40]{#2}}
\newcommand{\sanote}{\todo[color=violet!30]}
\newcommand{\note}{\todo[color=green!40]}
\newcommand{\newstart}{\note{The inserted text starts here}}
\newcommand{\newfinish}{\note{The inserted text finishes here}}
\setstcolor{red}
%***************************

@@

Both the Amplituhedron literature and the Wilson loop literature \hlfix{associate $n$ particle $N^kMHV$ interactions in SYM $N=4$}{make more prominent?} theory to volume forms on a set of positroid cells of $\Gr(k,n)$. This gives a CW complex of some submanifold in $\Gr(k,n)$, and understanding the geometry and topology of these spaces is the subject of ongoing research.

@@

There are a few things to note here.  First, what we call weakly admissible here is called admissible in (\cite{wilsonloop}, see definition 1.11 and section 1.3).  \hlfix{Many of the results below still apply when the hypothesis of admissibility is changed to weak admissibility.}{Propagate "weakly" through document}

Note also that if two propagators have the same pair of supporting edges, or if a propagator is supported on two adjacent edges, then the Wilson loop diagram is not admissible or even weakly admissible.

Further, if we take any admissible Wilson loop diagram and remove the non-supporting vertices, we obtain a weakly admissible Wilson loop diagram that may or may not be admissible itself.

For the remainder of the paper, we will restrict our attention to admissible Wilson loop diagrams and subdiagrams thereof.

@@

Matroid theory relates to the study of Wilson loop diagrams as follows. In \cite[section 3]{wilsonloop}, Agarwala and Marin-Amat show that every weakly admissible Wilson loop diagram $W$ with $k$ propagators defines a positroid $M(W)$ of rank $k$ on a base set of size $n$. It is a direct consequence of Postnikov's work, {\cite[Theorem 6.5]{Postnikov}},\hlfix{ that positroid cells are convex subset of $\Gr(k,n)$}{I  hate  saying it this way, but I can't find it directly written anywhere}. Putting these together gives that the Wilson loop diagrams  correspond to convex subsets of $\Gr(k,n)$, connecting the geometry of Wilson loop diagrams with the geometry underlying the Amplituhedron.

@@

\begin{thm} \label{thm WLD defines matroid} \cite[Theorem 3.6]{wilsonloop} Any Wilson loop diagram $W =(\cP, [n])$ defines a matroid $M(W)$ with ground set $[n]$. The independent sets are exactly those subsets $V \subseteq [n]$ \hlfix{such that for every subset of vertices, $U\subseteq V$, supports at least as many propagators as vertices in the set, $|\Prop(U)| \geq |U|$}{This edit isn't grammatically correct and is more complicated than what the referee asked for, could you fix it?}. \label{thm:WLDmatroid}\end{thm}
In other words, the independent sets of $M(W)$ correspond to the sets of vertices in $W$ with ``enough propagators sufficiently well distributed'' on them: not only should $V$ itself support at least as many propagators as the number of vertices it contains, but the same should be true of any subset of $V$ as well.

@@

\begin{lem}\label{lem facts about WLD matroids}
Let $W = (\cP,[n])$ be a weakly admissible Wilson loop diagram. Then:
\begin{enumerate}
\item The rank of a set $V \subseteq [n]$ is bounded above by $\min\{|V|,|\Prop(V)|\}$, with $\rk(V) = |V|$ if and only if $V$ is an independent set.
%\item Let $v \in [n]$ and $q \in \Prop(v)$. Then for any $V \subset [n]$ such that $q \not \in \Prop(V)$, we have $\rk(V \cup \{v\}) = \rk(V) +1$, i.e. adding a vertex that supports a new propagator to a set increases the rank of the set.
\item If $C \subseteq [n]$ is a cycle in $M(W)$, then $\rk(C) = |Prop(C)|$.
\item If $[n]$ can be partitioned into at least two non-empty sets, each of which support disjoint sets of propagators and in such a way that we obtain a partition of the propagator set, i.e. if we have \bas [n] = \sqcup_i V(P_i) \ \textrm{ such that } \ \sqcup P_i = \cP; \   \text{ with }V(P_i) \cap V(P_j) = \emptyset \text{ and } P_i \cap P_j = \emptyset \quad \forall i \neq j\;, \eas then the matroid $M(W)$ is separable, i.e. \bas M(W) = \bigoplus_i M(P_i, V(P_i)) \;.\eas 
\item The set $F(P)$ from Definition~\ref{VPropdfn} is a flat of $M( W)$, which we call the \emph{propagator flat} of $P$.
\item If $F$ is a cyclic flat of $M(W)$, then it is a propagator flat.
\end{enumerate}
\end{lem}
\begin{proof}
The first part of item 1 is \cite[Equation (9)]{wilsonloop} and surrounding discussion, and the second part is standard matroid theory. Item 2 is \cite[Lemma 3.27]{wilsonloop}. Item 3 is a direct consequence of \cite[Lemma 3.20]{wilsonloop}\sanote{ This is the right reference in both arXiv and CMP versions. Don't know what to do?} and the fact that $F(P_1)^c = V(P_1^c)$, and item 5 is given in \cite[Lemma 3.28]{wilsonloop}.

@@

\section{Matroids, triangulations, and equivalence classes of Wilson loop diagrams\label{sec equiv}} \todo{rephrase in terms of weakly.}

@@

There are many other instances of this basic methodology in combinatorics, but we'd like to point out one in particular that involves many objects closely related to those we use here, \cite{CGdyck}.\footnote{Thanks to the referee for the companion paper for pointing this out.}  In particular they, like us, look at a special case within the world of positroids and related objects which is enumerated by Catalan numbers.  For them these are positroids coming from unit interval graphs, while for us they are the equivalent exact subdiagrams.  Our objects are different, which can be seen by the fact that the exact subdiagram with 5 vertices (which is unique up to rotation) does not give a decorated permuation in the form of the Main Theorem of \cite{CGdyck}, but there are similarities in flavour with our arguments.

>>>>

%*****************
% Annotations
%\usepackage{soul}
\usepackage[colorinlistoftodos,textsize=footnotesize]{todonotes}
%\newcommand{\hlfix}[2]{\texthl{#1}\todo{#2}}
%\newcommand{\hlnew}[2]{\texthl{#1}\todo[color=green!40]{#2}}
%\newcommand{\sanote}{\todo[color=violet!30]}
%\newcommand{\note}{\todo[color=green!40]}
%\newcommand{\newstart}{\note{The inserted text starts here}}
%\newcommand{\newfinish}{\note{The inserted text finishes here}}
%\setstcolor{red}
%***************************

@@

Both the Amplituhedron literature and the Wilson loop literature associate $n$ particle $N^kMHV$ interactions in SYM $N=4$ theory to volume forms on a set of positroid cells of $\Gr(k,n)$. This gives a CW complex of some submanifold in $\Gr(k,n)$, and understanding the geometry and topology of these spaces is the subject of ongoing research.

@@

There are a few things to note here.  First, what we call weakly admissible here is called admissible in (\cite{wilsonloop}, see definition 1.11 and section 1.3).
Note also that if two propagators have the same pair of supporting edges, or if a propagator is supported on two adjacent edges, then the Wilson loop diagram is not admissible or even weakly admissible.
Further, if we take any admissible Wilson loop diagram and remove the non-supporting vertices, we obtain a weakly admissible Wilson loop diagram that may or may not be admissible itself.

For the remainder of the paper, we will restrict our attention to admissible and weakly admissible Wilson loop diagrams.

@@

Matroid theory relates to the study of Wilson loop diagrams as follows. In \cite[section 3]{wilsonloop}, Agarwala and Marin-Amat show that every weakly admissible Wilson loop diagram $W$ with $k$ propagators defines a positroid $M(W)$ of rank $k$ on a base set of size $n$. It is a direct consequence of Postnikov's work, {\cite[Theorem 6.5]{Postnikov}}, that positroid cells are convex subset of $\Gr(k,n)$. Putting these together gives that the Wilson loop diagrams  correspond to convex subsets of $\Gr(k,n)$, connecting the geometry of Wilson loop diagrams with the geometry underlying the Amplituhedron.

@@

\begin{thm} \label{thm WLD defines matroid} \cite[Theorem 3.6]{wilsonloop} Any Wilson loop diagram $W =(\cP, [n])$ defines a matroid $M(W)$ with ground set $[n]$. The independent sets are exactly those subsets $V \subseteq [n]$ such that every $U\subseteq V$ satisfies $|\Prop(U)| \geq |U|$. \label{thm:WLDmatroid}\end{thm}
In other words, the independent sets of $M(W)$ correspond to the sets of vertices in $W$ with ``enough propagators sufficiently well distributed'' on them: not only should $V$ itself support at least as many propagators as the number of vertices it contains, but the same should be true of any subset of $V$ as well.

@@

\begin{lem}\label{lem facts about WLD matroids}
Let $W = (\cP,[n])$ be a weakly admissible Wilson loop diagram. Then:
\begin{enumerate}
\item The rank of a set $V \subseteq [n]$ is bounded above by $\min\{|V|,|\Prop(V)|\}$, with $\rk(V) = |V|$ if and only if $V$ is an independent set.
%\item Let $v \in [n]$ and $q \in \Prop(v)$. Then for any $V \subset [n]$ such that $q \not \in \Prop(V)$, we have $\rk(V \cup \{v\}) = \rk(V) +1$, i.e. adding a vertex that supports a new propagator to a set increases the rank of the set.
\item If $C \subseteq [n]$ is a cycle in $M(W)$, then $\rk(C) = |Prop(C)|$.
\item If $[n]$ can be partitioned into at least two non-empty sets, each of which support disjoint sets of propagators and in such a way that we obtain a partition of the propagator set, i.e. if we have \bas [n] = \sqcup_i V(P_i) \ \textrm{ such that } \ \sqcup P_i = \cP; \   \text{ with }V(P_i) \cap V(P_j) = \emptyset \text{ and } P_i \cap P_j = \emptyset \quad \forall i \neq j\;, \eas then the matroid $M(W)$ is separable, i.e. \bas M(W) = \bigoplus_i M(P_i, V(P_i)) \;.\eas 
\item The set $F(P)$ from Definition~\ref{VPropdfn} is a flat of $M( W)$, which we call the \emph{propagator flat} of $P$.
\item If $F$ is a cyclic flat of $M(W)$, then it is a propagator flat.
\end{enumerate}
\end{lem}
\begin{proof}
The first part of item 1 is \cite[Equation (9)]{wilsonloop} and surrounding discussion, and the second part is standard matroid theory. Item 2 is \cite[Lemma 3.27]{wilsonloop}. Item 3 is a direct consequence of \cite[Theorem 3.20]{wilsonloop} and the fact that $F(P_1)^c = V(P_1^c)$, and item 5 is given in \cite[Lemma 3.28]{wilsonloop}.

@@

\section{Matroids, triangulations, and equivalence classes of Wilson loop diagrams\label{sec equiv}}

@@

There are many other instances of this basic methodology in combinatorics, but we'd like to point out one in particular that involves many objects closely related to those we use here, \cite{CGdyck}\footnote{Thanks to the referee for the companion paper, \cite{generalcombinatoricsII}, for pointing this out.}.  In particular they, like us, look at a special case within the world of positroids and related objects which is enumerated by Catalan numbers.  For them these are positroids coming from unit interval graphs, while for us they are the equivalent exact subdiagrams.  Our objects are different, which can be seen by the fact that the exact subdiagram with 5 vertices (which is unique up to rotation) does not give a decorated permuation in the form of the Main Theorem of \cite{CGdyck}, but there are similarities in flavour with our arguments.
